---
title: "Analysis of a thyroid gene expression dataset"
date: "January 20, 2019"
excerpt: "Exploratory analysis of a raw and normalized human thyroid gene expression dataset obtained from ARCHS4"
tags: "thyroid gene expression archs4 tsne"
toc: true
permalink: /archs4-thyroid-analysis/
---

```{r libraries}
# set options and load libraries
knitr::opts_chunk$set(echo = FALSE, fig.align="center", 
  message = FALSE, warning=FALSE, fig_width=10)

library(tidyverse)
library(ggplot2)
library(reshape2)
library(knitr)
library(pheatmap)
library(Rtsne)

# read the expression data (raw and normalized
expr <- read_tsv(snakemake@input$expr) %>%
  rename(gene_name = X1)
raw <- read_tsv(snakemake@input$raw) %>%
  rename(gene_name = X1)

# load design and do some cleaning
# NOTE: we only download the H. sapiens file
design <- read_tsv(snakemake@input$design) %>%
  mutate(tissue = trimws(tolower(gsub("Homo sapiens |Human","",tissue))))

# sanity check for sample ordering
print("Order of samples in expr and design is the same:")
all(colnames(expr %>% mutate(gene_name = NULL)) == design$sample)

```
# Inroduction

This is an exploratory analysis of the downloaded and processed ARCHS<sup>4</sup> 
gene expression data for a [thyroid](https://en.wikipedia.org/wiki/Thyroid) 
tissue specification. 
Data was obtained from ARCHS<sup>4</sup> using the non-official [ARCHS4 data loader](https://github.com/jhawe/archs4_loader)

## Gene expression data
The [central dogma of biology](https://en.wikipedia.org/wiki/Central_dogma_of_molecular_biology) 
states that genes on the DNA of a cell are transcribed into RNAs which are in turn further processed 
and translated into proteins.
Modern experimental technologies such as [next-generation sequencing](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC3841808/)
enable researchers to obtain a complete read-out of RNA molecules present in a specific 
biosample at the time of measurement. 
These read-outs can be quantified on a per gene (or per transcript) basis to obtain gene expression
data: A quantitative estimate of the amount of RNA molecules per gene present in individual biosamples.
Using these estimates, scientist seek to answer pressing biomedical questions such as 
to uncover the molecular mechanisms behind cancer or other diseases.

## ARCHS<sup>4<sup>
ARCHS<sup>4</sup> is a project which seeks to unite gene expression data form diverse sources
and provide easy access to a homogenously processed data matrix, containing hundreds of
individual samples, now ready to be jointly analyzed.
In this short analysis we'll have a look on ARCHS<sup>4</sup> Thyroid gene expression data
where we want to get an idea of how uniform these data are and whether additional
(manual) curation is necessary prior to downstream analysis.

## t-SNE
For our analysis we use the t-SNE algorithm.
TODO

# Data
In this analysis we will have have a look at the raw gene counts as well as the
normalized expression matrix as provided by the ARCHS<sup>4</sup> loader.

After loading the data, we see that we have a total of `r nrow(expr)` genes 
measured in `r ncol(expr)-1` samples (common for both raw and normalized data).
Below is a sample of the raw data matrix:

```{r load_data}

print("Expression:")
kable(expr[1:5,1:10])

```

In addition to the expression data, we the ARCHS<sup>4</sup> loader provides us with
a design table containing the following column names:

```{r design_cols}
colnames(design)
```

Before looking in detail at the expression data, we first have a look at the individual columns
of the design table in order to better get to know our data.
Below we show for each column of the design matrix the unique values contained
therein. This is a bit messy, but for now we accept this:

```{r values_by_cols}
lapply(colnames(design), function(x) {
  design %>% select(one_of(x)) %>% unique
})
```

Let's also check how many samples we get if we group by the individual values 
for each of the columns:

```{r samples_by_groups}
# we remove id, series and description
cols <- colnames(design)[-c(1,3,5)]

lapply(cols, function(x) {
  design %>% group_by_(.dots = as.symbol(x)) %>% 
    summarize(count=n()) %>% 
    arrange(desc(count))
})

```

TODO
describe what we sse

# Raw and batch normalized gene expression

Ok, now let's have a look at the actual gene expression data.
We will have a look at the histogram of all expression values and a heatmap
of the top 1% most variable genes. We will do this both for the normalized
data as well as the raw gene expression counts:

```{r expression_overview}
expr_subs <- expr
expr_subs$gene_name <- NULL
expr$gene_var <- apply(expr_subs,1,var)
expr_subs <- filter(expr, gene_var > quantile(gene_var, .99))

raw_subs <- raw
raw_subs$gene_name <- NULL
raw$gene_var <- apply(raw_subs,1,var)
raw_subs <- filter(raw, gene_var > quantile(gene_var, .99))

# histogram
toplot <- expr %>% mutate(gene_var = NULL, gene_name = NULL) %>%
  melt
ggplot(toplot, aes(x=value)) + geom_histogram() + xlab("expression") + 
  ggtitle("Distribution of expression values.")

# heatmap raw
raw_corr <- mutate(raw_subs, gene_var = NULL, gene_name = NULL) %>% t %>% cor
pheatmap(raw_corr, main = "Correlation map of top 1% variable genes (raw).")

# heatmap norm
expr_corr <- mutate(expr_subs, gene_var = NULL, gene_name = NULL) %>% t %>% cor
pheatmap(expr_corr, main = "Correlation map of top 1% variable genes (norm).")

```

TODO describe what we see
# t-SNE
Now let's do some tSNE plots to see whether we can see any specific clusters emerging.
Specifically, we will look at two types of t-SNE plots: One for the raw expression data as 
was extracted from ARCHS4 and one using the batch normalized data.

```{r tsne}

expr_subs <- t(mutate(expr, gene_var = NULL, gene_name = NULL))
raw_subs <- t(mutate(raw, gene_var = NULL, gene_name = NULL))

# tsne
reduction <- Rtsne(expr_subs, max_iter = 1000, theta = 0.0, 
                   dims = 2, perplexity = 30)
reduction_raw <- Rtsne(raw_subs, max_iter = 1000, theta = 0.0, 
                   dims = 2, perplexity = 30)

# plotting
toplot <- data.frame(rbind(reduction$Y, reduction_raw$Y))
colnames(toplot) <- c("dim1", "dim2")

toplot <- mutate(toplot, type = c(rep("normalized", nrow(reduction$Y)),
                                  rep("raw", nrow(reduction_raw$Y))))

toplot <- cbind(toplot, tissue = design$tissue, instrument = design$instrument,
                series=design$series)

ggplot(toplot, aes(x=dim1, y=dim2, col=tissue)) + 
  geom_point() + 
  facet_wrap(~type, nrow=2) +
  ggtitle("t-SNE on gene expression data labeled \nby 'tissue' meta-data information")

ggplot(toplot, aes(x=dim1, y=dim2, col=instrument)) + 
  geom_point() + 
  facet_wrap(~type, nrow=2) +
  ggtitle("t-SNE on gene expression data labeled \nby 'instrument' meta-data information")

ggplot(toplot, aes(x=dim1, y=dim2, col=series)) + 
  geom_point() + 
  facet_wrap(~type, nrow=2) +
  ggtitle("t-SNE on gene expression data labeled \nby 'Series' meta-data information")

```

TODO describe what we see
## Session Info
```{r}
sessionInfo()
```
