---
title: "Example of how to create a genome-wide genomic pairs plot"
date: "January 20, 2019"
excerpt: "This entry is about how to visualize genomic pairs data (e.g. eQTL, meQTL etc.) using tidyverse/ggplot2 and the GenomicRanges packages in R"
tags: "tidyverse ggplot2 QTL pairs matrix R GenomicRanges"
toc: true
permalink: /genomic-pairs-matrix/
---


```{r libraries, include=FALSE}
# set options and load libraries
knitr::opts_chunk$set(echo = FALSE, fig.align="center",
  message = FALSE, warning=FALSE, fig_width=10)

library(tidyverse)
library(reshape2)
library(GenomicRanges)
library(knitr)

```

# Inroduction

In this post, I want to give an example of how pairs of genomic entities, such
as genotypes and associated genes (QTLs), can be plotted in a genome-wide scale using
tidyverse/ggplot2 and the GenomicRanges package in R.
We will use publicly available trans-eQTL data (see below) and will create a
single overview matrix showing the frequencies of QTLs in these data.
To this end, we will bin/tile all standard chromosomes of the human reference genome
(GRCh37/hg19) into bins of 10MB, overlap these bins with our QTL results and then
plot them using a simple tile plot available in ggplot2.
In addition, we'll add margin plots to the main matrix-like plot, showing the 
row and column summary counts over all associations.

## Data
Expression quantitative trait loci (eQTL) are, in essence, statistical associations
between genetic variants/genotypes (e.g. single nucleotide polymorphisms, SNPs) and 
the expression of genes measured ideally in a large number of samples to obtain
sufficient statistical power.
In other words, eQTL are a way of pinpointing DNA sequence variants which, most likely, 
have a functional impact in the many complex mechanisms taking place in the studied organism.
Generally, eQTL can be classified in cis- and trans-eQTL, cis meaning that the SNP
and the respective gene reside on the same chromsome and trans meaning that the two
entities are located on different chromosomes.
For our example, we will focus on trans-eQTL only.
We downloaded all significant trans-eQTL data identified by the [eQTLgen consortium](https://eqtlgen.org/trans-eqtls.html) and use those to generate a nice
overview on trans-eQTLs identified in whole blood data.


# Hands-on

## Data preparation
As a first step, we load the eQTLgen trans-eQTLs and create GenomicRanges objects, i.e.
objects to easily handle genomic position annotations.

```{r echo=T}
# load eQTL data
eqtl <- read_tsv("https://molgenis26.gcc.rug.nl/downloads/eqtlgen/trans-eqtl/trans-eQTL_significant_20181017.txt.gz")

# get the number of associations per SNP
ntrans_per_snp <- group_by(eqtl, SNP) %>% summarise(ntrans=n())

# filter hotspots, i.e. retain only SNPs with at least 5 trans associations
eqtl <- mutate(eqtl, ntrans = ntrans_per_snp[match(SNP, ntrans_per_snp$SNP),]$ntrans) %>%
  filter(ntrans >= 5)

# get the hg19 chromosome definitions
library(BSgenome.Hsapiens.UCSC.hg19)
hg19info <- seqinfo(BSgenome.Hsapiens.UCSC.hg19)

# create the ranges objects
eqtl_ranges <- with(
  eqtl,
  GRanges(
    paste0("chr", SNPChr),
    IRanges(SNPPos, width = 1),
    name = SNP,
    trans_associations = ntrans,
    seqinfo = hg19info
  )
)

trans_genes <- with(eqtl,
                    GRanges(
                      paste0("chr", GeneChr),
                      IRanges(GenePos, width = 2),
                      name = GeneSymbol,
                      seqinfo = hg19info
                    ))

```

Ok, nice! We now got all our data loaded and have them neatly available as GenomicRanges objects.
We are looking at `r print(length(unique(eqtl_ranges)))` SNPs which are associated with a total of `r length(unique(trans_genes))` genes (`r print(length(eqtl_ranges))` total associations).
Now, since we want to do a genome-wide plot, we need to get some information on the size of the chromosomes etc. in order to be able to indicate chromosome boundaries.
For that, we get the hg19 genome annotation and extract the sequence lengths for chromosomes 1-22.
We further tile our genome information into tiles (or bins) of size 10MB. 
These tiles will be used later on to map the individual genomic positions from the SNPs and genes to the
respective position in the plot.

```{r echo=T}
# get chromoeome lengths
chrs = paste0("chr", 1:22)
chrlen <- seqlengths(hg19info)
chrlen <- chrlen[paste0("chr", 1:22)]

# tile the genome
genome_bins <-
  tileGenome(chrlen,
             tilewidth = 1e7,
             cut.last.tile.in.chrom = T)

# define breaks (used to get a nice, scaled 
# grid visualization)
breaks <- table(seqnames(genome_bins))
for (i in 2:length(breaks)) {
  breaks[i] <- breaks[i - 1] + breaks[i]
}
```

With the above steps done, we can now almost start creating the plot.
Last remaining things to do is to map our eQTL entities to the respective genome bins and to create the summary
counts for the x and y (row and column) margins.
Let's do that now:

```{r echo=T}
# get overlaps 
bin_overlaps_eqtl <- findOverlaps(eqtl_ranges, genome_bins)
bin_overlaps_genes <- findOverlaps(trans_genes, genome_bins)

# to be save, we only retain pairs where both entities could be maped
# to chromosomes 1:22
mappable_pairs <-
  intersect(queryHits(bin_overlaps_eqtl), queryHits(bin_overlaps_genes))
x_bin <-
  subjectHits(bin_overlaps_eqtl)[queryHits(bin_overlaps_eqtl) %in% mappable_pairs]
y_bin <-
  subjectHits(bin_overlaps_genes)[queryHits(bin_overlaps_genes) %in% mappable_pairs]

# create a data.frame for plotting with ggplot2
pairs_binned <- cbind.data.frame(x_bin, y_bin)
pairs_binned <-
  pairs_binned[order(pairs_binned$x_bin, pairs_binned$y_bin),]

# the margins, i.e. the individual totals for the x and y bins
x_margin <- group_by(pairs_binned, x_bin) %>% summarise(count = n())
y_margin <- group_by(pairs_binned, y_bin) %>% summarise(count = n())

```

Nice, now everything is prepared and we can start with the actual plotting!

## The actual plotting
Let's keep it brief, I added some comments to the code (as I'm supposed to do anyway) so you can figure out what is happening, but I kept it in a single code block so you can just copy, paste and adjust if you want.

```{r echo=T}
# set a less obstrusive theme and adjust margins
theme_set(theme_bw())
theme_update(plot.margin = margin(0, 0, 0, 0, "cm"))

# create the plot for the x-axis margin, single points
xmp <-
  ggplot(x_margin) + geom_point(aes(x = x_bin, y = count, size=count)) + 
  scale_size_continuous(guide=F) + 
  scale_x_continuous(
    expand = c(0.01, 0.01),
    breaks = as.vector(breaks),
    labels = NULL
  ) + xlab("") + theme(plot.margin = margin(0.1, 0, 0, 0.1, "cm"))

# create the plot for the y-axis margin, single points
ymp <- ggplot(y_margin) +
  geom_point(aes(x = y_bin, y = count, size=count)) +
  coord_flip() + scale_x_continuous(
    expand = c(0.01, 0.01),
    breaks = as.vector(breaks),
    labels = NULL
  ) + xlab("") + theme(
    plot.margin = margin(0, 0.1, 0.1, 0, "cm"),
    axis.text.x = element_text(
      angle = 90,
      vjust = 0.5,
      hjust = 1
    )
  ) + 
  scale_size_continuous(guide=F)

# the main matrix plot showing the eQTL information
g <-
  ggplot(pairs_binned) +
  geom_tile(fill = "red", aes(x = x_bin, y = y_bin)) +
  theme(
    text = element_text(size = 11),
    legend.text = element_text(size = 8),
    axis.text.x = element_text(vjust = 0.5, angle = 90),
    legend.title = element_text(size = 10),
    plot.margin = margin(0, 0, 0, 0, "cm")
  ) +
  xlab("SNPs") + ylab("genes") +
  scale_x_continuous(
    expand = c(0.01, 0.01),
    breaks = as.vector(breaks),
    labels = names(breaks)
  ) +
  scale_y_continuous(
    expand = c(0.01, 0.01),
    breaks = as.vector(breaks),
    labels = names(breaks)
  )

# now we got the individual plots, arrange them in a single
# large plot using grid.arrange
wr <- c(85, 15)
hr <- c(15, 85)

grid.arrange(xmp, nullGrob(), g, ymp,
             ncol = 2, widths = wr, heights = hr)
```

So, as you can see, we create our final plot by arrange three distinct plots in a single frame using the *grid.arrange* function from the *gridExtra* package.
This actually looks rather good! So good in fact that, we are content for now and stop tweaking the plot (I'm lazy today). 
Anyway, we can see that there are some SNPs or LD blocks which exhibit a huge number of trans associations as well as some regions in the genome which harbor a relatively larger number of genes which are influenced by these SNPs in trans as compared to other regions.


## Conclusion
Alright, you have seen how we can get an impression of genome-wide results from an eQTL study, just by using tidyverse/ggplot2 and the GenomicRanges packages in R. 
Of course, there are different ways of doing this, for example you could try to do a circos plot using the [ggbio](https://bioconductor.org/packages/release/bioc/html/ggbio.html) package in R or you can add continuous information about the number of eQTLs per bin directly in the main matrix plot (e.g. using the *fill* aesthetics).
Maybe I'll extend the plot later to show how to do this, but in the meanwhile: have fun plaiyng around with this example!

Until then, farewell!

# Session Info
```{r}
sessionInfo()
```



```{r include=F, render=F, eval=F}
# TODO we could add an example for continuoius value plotting
# continuous plot
pairs_binned <-
data.frame(table(pairs_binned), stringsAsFactors = F)
pairs_binned[, 1] <- as.numeric(as.character(pairs_binned[, 1]))
pairs_binned[, 2] <- as.numeric(as.character(pairs_binned[, 2]))
pairs_binned[pairs_binned$Freq > 50, "Freq"] <- 50


g <-
  ggplot(pairs_binned) +
  geom_tile(aes(x = x_bin, y = y_bin, fill = Freq)) +
  theme(
    text = element_text(size = 11),
    legend.text = element_text(size = 8),
    axis.text.x = element_text(vjust = 0.5, angle = 90),
    legend.title = element_text(size = 10),
    plot.margin = margin(0, 0, 0, 0, "cm"),
    legend.position = "bottom",
    legend.box.margin = margin(0.01, 0.01, 0.01, 0.01, "cm")
  ) +
  xlab(label_x) + ylab(label_y) +
  scale_x_continuous(
    expand = c(0.01, 0.01),
    breaks = as.vector(breaks),
    labels = names(breaks)
  ) +
  scale_y_continuous(
    expand = c(0.01, 0.01),
    breaks = as.vector(breaks),
    labels = names(breaks)
  ) +
  scale_fill_continuous(low = 'white',
                        high = 'red',
                        na.value = 'white')

gg <- ggplotGrob(g)$grobs
legend <-
  gg[[which(sapply(gg, function(x)
    x$name) == "guide-box")]]

#lheight <- sum(legend$height)
#lwidth <- sum(legend$width)
g <- g + theme(legend.position = "none")

# the discretized plot


```
